\section{Chomsky Normal Form}
\label{ChomskyNormalForm}

In this section, we study a special form of grammars called Chomsky
Normal Form (CNF), which was named after the linguist Noam Chomsky.
Grammars in CNF have very nice formal properties.  In particular,
valid parse trees for grammars in CNF are very close to being binary
trees.

Any grammar that doesn't generate $\%$ can be put in CNF.  And, if $G$
is a grammar that does generate $\%$, it can be turned into a grammar
in CNF that generates $L(G)-\{\%\}$.  In the next section, we will use
this fact when proving the pumping lemma for context-free languages, a
method for showing the certain languages are not context-free.

We will begin by giving an algorithm for turning a grammar $G$ into a
simplified grammar with no productions of the form $q\fun\%$ and
$q\fun r$. This will enable us to give an algorithm that takes in a
grammar $G$, and calculates $L(G)$, when it is finite, and reports
that it is infinite, otherwise.

\subsection{Removing $\%$-Productions}

\index{grammar!removing $\%$-productions}%
Recall that an $\%$-\emph{production} is a production of the form
$q\fun\%$.  We will show by example how to turn a grammar $G$ into a
simplified grammar with no $\%$-productions that generates
$L(G)-\{\%\}$.

Suppose $G$ is the grammar
\begin{align*}
\Asf &\fun \mathsf{0A1} \mid \mathsf{BB} , \\
\Bsf &\fun \% \mid \mathsf{2B} .
\end{align*}

First, we determine which variables $q$ are \emph{nullable} in the
sense that $\%\in\Pi_q$, i.e., that $\%$ is the yield of a valid parse
tree for $G$ whose root label is $q$.
\begin{itemize}
\item Clearly, $\Bsf$ is nullable.

\item Since $\Asf\fun\Bsf\Bsf\in P_G$, it follows that $\Asf$ is
  nullable.
\end{itemize}

Now we use this information to compute the productions of
our new grammar.
\begin{itemize}
\item Since $\Asf$ is nullable, we replace the production
  $\Asf\fun\mathsf{0A1}$ with the productions $\Asf\fun\mathsf{0A1}$
  and $\Asf\fun\mathsf{01}$.  The idea is that this second production
  will make up for the fact that $\Asf$ won't be nullable in the new
  grammar.

\item Since $\Bsf$ is nullable, we replace the production
  $\Asf\fun\mathsf{BB}$ with the productions $\Asf\fun\mathsf{BB}$ and
  $\Asf\fun\mathsf{B}$ (the result of deleting either one of the
  $\Bsf$'s).

\item The production $\Bsf\fun\%$ is deleted.

\item Since $\Bsf$ is nullable, we replace the production
  $\Bsf\fun\mathsf{2B}$ with the productions $\Bsf\fun\mathsf{2B}$ and
  $\Bsf\fun\mathsf{2}$.
\end{itemize}
(If a production has $n$ occurrences of nullable variables in its
right side, then there will be $2^n$ new right sides, corresponding to
all ways of deleting or not deleting those $n$ variable
occurrences. But if a right side of $\%$ would result, we don't
include it.)
This give us the grammar
\begin{align*}
\Asf &\fun \mathsf{0A1} \mid \mathsf{01} \mid \mathsf{BB} \mid \Bsf , \\
\Bsf &\fun \mathsf{2B} \mid \twosf .
\end{align*}
In general, we finish by simplifying our new grammar.  The new grammar
of our example is already simplified, however.

\subsection{Removing Unit Productions}

\index{grammar!removing unit productions}
Recall that a \emph{unit production} for a grammar $G$ is a production
of the form $q\fun r$, where $r$ is a variable (possibly equal to
$q$).  We now show by example how to turn a grammar $G$ into a
simplified grammar with no $\%$-productions or unit productions that
generates $L(G)-\{\%\}$.

Suppose $G$ is the grammar
\begin{align*}
\Asf &\fun \mathsf{0A1} \mid \mathsf{01} \mid \mathsf{BB} \mid \Bsf , \\
\Bsf &\fun \mathsf{2B} \mid \twosf .
\end{align*}
We begin by applying our algorithm for removing $\%$-productions to
our grammar; the algorithm has no effect in this case.

Our new grammar will have the same variables and start variable as
$G$.  Its set of productions is the set of all $q\fun w$ such that
$q$ is a variable of $G$, $w\in\Str$ doesn't consist of a single variable
of $G$, and there is a variable $r$ such that
\begin{itemize}
\item $r$ is parsable from $q$, and

\item $r\fun w$ is a production of $G$.
\end{itemize}
(Determining whether $r$ is parsable from $q$ is easy, since we are
working with a grammar with no $\%$-productions.)

This process results in the grammar
\begin{align*}
\Asf &\fun \mathsf{0A1} \mid \mathsf{01} \mid \mathsf{BB} \mid
\mathsf{2B} \mid \twosf , \\
\Bsf &\fun \mathsf{2B} \mid \twosf .
\end{align*}
Finally, we simplify our grammar, which gets rid of the production
$\Asf\fun\mathsf{2B}$, giving us the grammar
\begin{align*}
\Asf &\fun \mathsf{0A1} \mid \mathsf{01} \mid \mathsf{BB} \mid \twosf , \\
\Bsf &\fun \mathsf{2B} \mid \twosf .
\end{align*}

\subsection*{Removing $\%$ and Unit Productions in Forlan}

The Forlan module \texttt{Gram} defines the following functions:
\begin{verbatim}
val eliminateEmptyProductions        : gram -> gram
val eliminateEmptyAndUnitProductions : gram -> gram
\end{verbatim}
\index{Gram@\texttt{Gram}!eliminateEmptyProductions@\texttt{eliminateEmptyProductions}}%
\index{Gram@\texttt{Gram}!eliminateEmptyAndUnitProductions@\texttt{eliminateEmptyAndUnitProductions}}%

For example, if \texttt{gram} is the grammar
\begin{align*}
\Asf &\fun \mathsf{0A1} \mid \mathsf{BB} , \\
\Bsf &\fun \% \mid \mathsf{2B} .
\end{align*}
then we can proceed as follows.
\input{chap-4.9-forlan1}

\subsection{Generating a Grammar's Language When Finite}

\index{grammar!generating language when finite}%
We can now give an algorithm that takes in a grammar $G$ and generates
$L(G)$, when it is finite, and reports that $L(G)$ is infinite,
otherwise. The algorithm begins by letting $G'$ be the result of
eliminating $\%$-productions and unit productions from $G$. Thus $G'$
is simplified and generates $L(G)-\{\%\}$.

If there is recursion in the productions of $G'$---either direct or
mutual---then there is a variable $q$ of $G'$ and a valid parse tree
$\pt$ for $G'$, such that the height of $\pt$ is at least one, $q$ is
the root label of $\pt$, and the yield of $\pt$ has the form $xqy$,
for strings $x$ and $y$, each of whose symbols is in
$\alphabet\,G'\cup Q_{G'}$. (In particular, $q$ may appear in $x$ or
$y$.) Because $G'$ lacks $\%$-productions and unit productions, it
follows that $x\neq\%$ or $y\neq\%$. Because each variable of $G'$ is
generating, we can turn $\pt$ into a valid parse tree $\pt'$ whose
root label is $q$, and whose yield has the form $uqv$, for
$u,v\in(\alphabet\,G')^*$, where $u\neq\%$ or $v\neq\%$.

Thus we have that $uqv$ is parsable from $q$ in $G$', and an easy
mathematical induction shows that $u^nqv^n$ is parsable from $q$ in
$G'$, for all $n\in\nats$.  Because $u\neq\%$ or $v\neq\%$, and $q$ is
generating, it follows that there are infinitely many strings
generated from $q$ in $G'$. And, since $q$ is reachable, and every
variable of $G'$ is generating, it follows that $L(G')$, and thus
$L(G)$, is infinite.

Consequently, our algorithm can continue as follows. If the
productions of $G'$ have recursion, then it reports that $L(G)$ is
infinite.  Otherwise, it calculates $L(G')$ from the bottom-up, and
adds $\%$ iff $G$ generates $\%$.

The Forlan module \texttt{Gram} defines the following function:
\begin{verbatim}
val toStrSet : gram -> str set
\end{verbatim}
\index{Gram@\texttt{Gram}!toStrSet@\texttt{toStrSet}}%
Suppose \texttt{gram} is the grammar
\begin{align*}
\Asf &\fun \mathsf{BB}, \\
\Bsf &\fun \mathsf{CC}, \\
\Csf &\fun \% \mid \zerosf \mid \onesf ,
\end{align*}
and
\texttt{gram'} is the grammar
\begin{align*}
\Asf &\fun \mathsf{BB}, \\
\Bsf &\fun \mathsf{CC}, \\
\Csf &\fun \% \mid \zerosf \mid \onesf \mid \Asf .
\end{align*}
Then we can proceed as follows:
\input{chap-4.9-forlan3}  

Suppose we have a grammar $G$ and a natural number $n$, and we wish to
generate the set of all elements of $L(G)$ of length $n$.  We can start by
creating an EFA $M$ accepting all strings over the alphabet of $G$
with length $n$.  Then, we can intersect $G$ with $M$, and apply
\texttt{Gram.toStrSet} to the resulting grammar.

\subsection{Chomsky Normal Form}

\index{grammar!Chomsky Normal Form}%
\index{Chomsky Normal Form}%
A grammar $G$ is in \emph{Chomsky Normal Form} (CNF) iff
each of its productions has one of the following forms:
\begin{itemize}
\item $q\fun a$, where $a$ is not a variable; and

\item $q\fun pr$, where $p$ and $r$ are variables.
\end{itemize}

We explain by example how a grammar $G$ can be turned into a
simplified grammar in CNF that generates $L(G)-\{\%\}$.
Suppose $G$ is the grammar
\begin{align*}
\Asf &\fun \mathsf{0A1} \mid \mathsf{01} \mid \mathsf{BB} \mid \twosf , \\
\Bsf &\fun \mathsf{2B} \mid \twosf .
\end{align*}
\begin{itemize}
\item We begin by applying our algorithm for removing $\%$-productions
  and unit productions to this grammar.  In this case, it has no
  effect.

\item Since the productions $\Asf\fun\mathsf{BB}$, $\Asf\fun\twosf$
  and $\Bsf\fun\twosf$ are legal CNF productions, we simply transfer
  them to our new grammar.

\item Next we add the variables $\Zero$, $\One$ and $\Two$ to our
  grammar, along with the productions
  \begin{gather*}
    \Zero\fun\zerosf,\quad\One\fun\onesf,\quad\Two\fun\twosf .
  \end{gather*}

\item Now, we can replace the production $\Asf\fun\mathsf{01}$ with
  $\Asf\fun\Zero\One$.  And, we can replace the production
  $\Bsf\fun\mathsf{2B}$ with the production $\Bsf\fun\Two\Bsf$.

\item Finally, we replace the production $\Asf\fun\mathsf{0A1}$ with
  the productions
  \begin{gather*}
    \Asf\fun\Zero\Csf,\quad\Csf\fun\Asf\One ,
  \end{gather*}
  and add $\Csf$ to the set of variables of our new grammar.
\end{itemize}
Summarizing, our new grammar is
\begin{align*}
  \Asf &\fun \mathsf{BB} \mid \twosf \mid \Zero\One \mid
  \Zero\Csf , \\
  \Bsf &\fun \twosf \mid \Two\Bsf , \\
  \Zero &\fun \zerosf , \\
  \One &\fun \onesf , \\
  \Two &\fun \twosf , \\
  \Csf &\fun \Asf\One .
\end{align*}

The official version of our algorithm names variables in a different
way.

\subsection*{Converting to Chomsky Normal Form in Forlan}

The Forlan module \texttt{Gram} defines the following function:
\begin{verbatim}
val chomskyNormalForm : gram -> gram
\end{verbatim}
\index{Gram@\texttt{Gram}!chomskyNormalForm@\texttt{chomskyNormalForm}}%

Suppose \texttt{gram} of type \texttt{gram} is bound to the grammar with
variables $\Asf$ and $\Bsf$, start variable $\Asf$, and productions
\begin{align*}
  \Asf &\fun \mathsf{0A1} \mid \mathsf{BB} , \\
  \Bsf &\fun \% \mid \mathsf{2B} .
\end{align*}
Here is how Forlan can be used to turn this grammar into a CNF
grammar that generates the nonempty strings that are generated by
\texttt{gram}:
\input{chap-4.9-forlan5}

\subsection{Notes}

The material in this section is standard.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "book"
%%% End: 
