\chapter{Recursive and Recursively Enumerable
Languages}
\label{RecursiveAndRecursivelyEnumerableLanguages}

In this chapter, we will study a universal programming language, which
we will use to define the recursive and recursively enumerable
languages.  We will see that the context-free languages are a proper
subset of the recursive languages, that the recursive languages are a
proper subset of the recursively enumerable languages, and that there
are languages that are not recursively enumerable.  Furthermore, we
will learn that there are problems, like the halting problem (the
problem of determining whether a program halts when run on a given
input), or the problem of determining if two grammars generate the
same language, that can't be solved by programs.

Traditionally, one uses Turing machines for the universal programming
language.  Turing machines are finite automata that manipulate
infinite tapes.  Although Turing machines are very appealing in some
ways, they are rather far-removed from conventional programming
languages, and are hard to build and reason about.

Instead, we will work with a simple functional programming language.
This language will have the same power as Turing machines, but will be
much easier to program in and reason about than Turing machines.
An ``implementation'' of our language (or of Turing machines) on a
real computer will run out of resources on some programs.

\input{chap-5.1}
\input{chap-5.2}
\input{chap-5.3}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "book"
%%% End: 
